Reconstructing Propositional Proofs in Type Theory

### 2017-10-25 paper/sections/proof-reconstruction/resolve.tex
more fixes.

In the last example in (\ref{ex:reorder-example}), we could not build the conjunction
$r ∧ (q ∨ p)$ since $p ∨ q$ is not syntactical equal to $q ∨ p$.
We solve this issue in the Lemma~\ref{lem:reorder-and-or} by
using the $\fconjunct_{∨}$ function defined in~(\ref{eq:conjunct-or}).

\begin{mainlemma}
  \label{lem:conjunct-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \fconjunct_{∨}~φ~ψ$ where,
  \begin{equation*}
    \begin{split}
    \label{eq:conjunct-or}
      &\fconjunct_{∨} : \Prop → \Prop \to \Prop\\
      &\fconjunct_{∨}~φ~ψ =
      \begin{cases}
        ψ, &\text{ if }φ ≡ ψ\\
        ψ, &\text{ if } ψ ≡ \freorder_{∨}~φ~ψ\\
        ψ, &\text{ if }ψ ≡ ψ₁ ∧ ψ₂,\ ψ₁ ≡ \fconjunct_{∨}~φ~ψ₁,\\
           &\text{ and } ψ₂ ≡ \freorder_{∨}~φ~ψ₂\\
        ψ, &\text{ if }φ ≡ φ₁ ∧ φ₂,\ ψ ≡ \fconjunct_{∨}~φ₁~ψ \\
        ψ, &\text{ if }φ ≡ φ₁ ∧ φ₂,\ ψ ≡ \fconjunct_{∨}~φ₂~ψ \\
        φ, &\text{ otherwise.}
      \end{cases}
    \end{split}
  \end{equation*}
\end{mainlemma}

@@

Now, we are ready to reconstruct the \resolve rule. This is the
first application of the Lemma \ref{lem:reorder-and-or}.
The \resolve rule is a generalization of the resolution theorem.
This rule has three arguments: a positive literal $ℓ$to apply the
resolution theorem and two derivations.
Recall a\emph{literal} is an atom (positive literal) or a negation
of an atom (negative literal).
The positive literal $ℓ$ must occur in
the formula from the first derivation and the
negative literal $¬ ℓ$ must occur in the formula from the second derivation,
see the pattern of \emph{resolve} rule in Fig.~\ref{fig:metis-inferences}.

>>>>

In the last example in (\ref{ex:reorder-example}), we could not build the conjunction
$r ∧ (q ∨ p)$ since $p ∨ q$ is not syntactical equal to $q ∨ p$.
We solve this issue in Lemma~\ref{lem:reorder-and-or} by
using the $\fconjunct_{∨}$ function defined in~(\ref{eq:conjunct-or}).

\begin{mainlemma}
  \label{lem:conjunct-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \fconjunct_{∨}~φ~ψ$ where,
  \begin{equation}
    \begin{split}
    \label{eq:conjunct-or}
      &\fconjunct_{∨} : \Prop → \Prop \to \Prop\\
      &\fconjunct_{∨}~φ~ψ =
      \begin{cases}
        ψ, &\text{ if }φ ≡ ψ\\
        ψ, &\text{ if } ψ ≡ \freorder_{∨}~φ~ψ\\
        ψ, &\text{ if }ψ ≡ ψ₁ ∧ ψ₂,\ ψ₁ ≡ \fconjunct_{∨}~φ~ψ₁,\\
           &\text{ and } ψ₂ ≡ \freorder_{∨}~φ~ψ₂\\
        ψ, &\text{ if }φ ≡ φ₁ ∧ φ₂,\ ψ ≡ \fconjunct_{∨}~φ₁~ψ \\
        ψ, &\text{ if }φ ≡ φ₁ ∧ φ₂,\ ψ ≡ \fconjunct_{∨}~φ₂~ψ \\
        φ, &\text{ otherwise.}
      \end{cases}
    \end{split}
  \end{equation}
\end{mainlemma}

@@

Now, we are ready to reconstruct the \resolve rule. This is the
first application of Lemma \ref{lem:reorder-and-or}.
The \resolve rule is a generalization of the resolution theorem.
This rule has three arguments: a positive literal $ℓ$to apply the
resolution theorem and two derivations.
Recall a\emph{literal} is an atom (positive literal) or a negation
of an atom (negative literal).
The positive literal $ℓ$ must occur in
the formula from the first derivation and the
negative literal $¬ ℓ$ must occur in the formula from the second derivation,
see the pattern of \emph{resolve} rule in Fig.~\ref{fig:metis-inferences}.

### 2017-10-25 paper/sections/proof-reconstruction/strip.tex
more fixes.

In this example, the conjecture \verb!(p & r) & q! is stripped into
tree subgoals: $p$, $p ⇒ r$ and $(p ∧ r) ⇒ q$.

@@

\begin{equation}
\label{eq:strip-fixed}
\begin{aligned}
&\hspace{.495mm}\fstrip₁ : \Prop → \Nat → \Prop\\
&\begin{array}{llll}
\fstrip₁ &(φ₁ ∧ φ₂)     &(\suc~n) &= \fuh~(\fstrip₁~φ₁~n) ∧ \fuh~(φ₁ ⇒ \fstrip₁~φ₂~n)\\
\fstrip₁ &(φ₁ ∨ φ₂)     &(\suc~n) &= \fuh~(¬ φ₁ ⇒ \fstrip₁~φ₂~n)\\
\fstrip₁ &(φ₁ ⇒ φ₂)     &(\suc~n) &= \fuh~(φ₁ ⇒ \fstrip₁~φ₂~n)\\
\fstrip₁ &(¬ (φ₁ ∧ φ₂)) &(\suc~n) &= \fuh~(φ₁ ⇒ \fstrip₁~¬ φ₂~n)\\
\fstrip₁ &(¬ (φ₁ ∨ φ₂)) &(\suc~n) &= \fuh~(\fstrip₁~¬ φ₁~n) ∧ \fuh~(¬ φ₁ ⇒ \fstrip₁~¬ φ₂~n)\\
\fstrip₁ &(¬ (φ₁ ⇒ φ₂)) &(\suc~n) &= \fuh~(\fstrip₁~φ₁~n) ∧ \fuh~(φ₁ ⇒ \fstrip₁~¬ φ₂~n)\\
\fstrip₁ &(¬ ¬ φ₁)      &(\suc~n) &= \fuh~(\fstrip₁~φ₁~n)\\
\fstrip₁ &(¬ ⊥)         &(\suc~n) &= ⊤\\
\fstrip₁ &(¬ ⊤)         &(\suc~n) &= ⊥\\
\fstrip₁ &φ             &\text{n} &= φ
\end{array}
\end{aligned}
\end{equation}

@@

$\bullet$~Case $φ ≡ φ₁ ∧ φ₂$. We prove $Γ ⊢ φ₁$ and $Γ ⊢ φ₂$. Using the
introduction rule for conjunction, the expected result follows.

@@

$\bullet$~Case $φ ≡ ¬ (φ₁ ∨ φ₂)$.
We show the theorems $Γ ⊢ ¬ φ₁$ and $Γ ⊢ ¬ φ₂$.
With the conjunction of $¬ φ₁$ and $¬ φ₂$ by applying De Morgan Law, the
result follows.

\begin{equation*}
(\mathcal{D})\hspace{3mm}
\begin{bprooftree}
\AxiomC{$Γ ⊢ \fstrip₁~(¬ (φ₁ ∨ φ₂))~(\suc~n)$}
\RightLabel{by~(\ref{eq:strip-fixed})}
\UnaryInfC{$Γ ⊢ \fuh~(\fstrip₁~¬ φ₁~n) ∧ \fuh~(¬ φ₁ ⇒ \fstrip₁~¬ φ₂~n)$}
\RightLabel{∧-proj₁}
\UnaryInfC{$Γ ⊢ \fuh~(\fstrip₁~¬ φ₁~n)$}
\RightLabel{Lemma~\ref{lem:inv-uh-lem}}
\UnaryInfC{$Γ ⊢ \fstrip₁~¬ φ₁~n$}
\RightLabel{by~ind.~hyp.}
  \UnaryInfC{$Γ ⊢ ¬φ₁$}
\end{bprooftree}
\end{equation*}
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$\mathcal{D}$}
  \UnaryInfC{$Γ ⊢ ¬ φ₁$}
  \AxiomC{$Γ ⊢ \fstrip₁~(¬ (φ₁ ∨ φ₂))~(\suc~n)$}
  \RightLabel{by~(\ref{eq:strip-fixed})}
  \UnaryInfC{$Γ ⊢ \fuh~(\fstrip₁~¬ φ₁~n) ∧ \fuh~(¬ φ₁ ⇒ \fstrip₁~¬ φ₂~n)$}
  \RightLabel{∧-proj₂}
  \UnaryInfC{$Γ ⊢ \fuh~(¬ φ₁ ⇒ \fstrip₁~¬ φ₂~n)$}
  \RightLabel{Lemma~\ref{lem:inv-uh-lem}}
  \UnaryInfC{$Γ ⊢ ¬ φ₁ ⇒ \fstrip₁~¬ φ₂~n$}
  \RightLabel{⇒-elim}
  \BinaryInfC{$Γ ⊢ \fstrip₁~¬ φ₂~n$}
  \RightLabel{by~ind.~hyp.}
  \UnaryInfC{$Γ ⊢ ¬ φ₂$}
  \end{bprooftree}
\end{equation*}
\end{proof} % the proof continue...

The following theorem is convenient to substitute equals by equals in
a theorem. Recall the equality ≡ is symmetric and transitive as well.
We use such properties without any mention.

>>>>

In this example, the conjecture \$(p ∧ r) ∧ q$ is stripped into
tree subgoals: \$p$, \$p ⇒ r$ and \$(p ∧ r) ⇒ q$.

@@

\begin{equation}
\label{eq:strip-fixed}
\begin{aligned}
&\hspace{.495mm}\fstrip₁ : \Prop → \Nat → \Prop\\
&\begin{array}{llll}
\fstrip₁ &(φ₁ ∧ φ₂)     &(\suc~n) &= \fuh~(\fstrip₁~φ₁~n) ∧ \fuh~(φ₁ ⇒ \fstrip₁~φ₂~n)\\
\fstrip₁ &(φ₁ ∨ φ₂)     &(\suc~n) &= \fuh~(¬ φ₁ ⇒ \fstrip₁~φ₂~n)\\
\fstrip₁ &(φ₁ ⇒ φ₂)     &(\suc~n) &= \fuh~(φ₁ ⇒ \fstrip₁~φ₂~n)\\
\fstrip₁ &(¬ (φ₁ ∧ φ₂)) &(\suc~n) &= \fuh~(φ₁ ⇒ \fstrip₁~(¬ φ₂)~n)\\
\fstrip₁ &(¬ (φ₁ ∨ φ₂)) &(\suc~n) &= \fuh~(\fstrip₁~(¬ φ₁)~n) ∧ \fuh~(¬ φ₁ ⇒ \fstrip₁~(¬ φ₂)~n)\\
\fstrip₁ &(¬ (φ₁ ⇒ φ₂)) &(\suc~n) &= \fuh~(\fstrip₁~φ₁~n) ∧ \fuh~(φ₁ ⇒ \fstrip₁~(¬ φ₂)~n)\\
\fstrip₁ &(¬ ¬ φ₁)      &(\suc~n) &= \fuh~(\fstrip₁~φ₁~n)\\
\fstrip₁ &(¬ ⊥)         &(\suc~n) &= ⊤\\
\fstrip₁ &(¬ ⊤)         &(\suc~n) &= ⊥\\
\fstrip₁ &φ             &\text{n} &= φ
\end{array}
\end{aligned}
\end{equation}

@@

$\bullet$~Case $φ ≡ φ₁ ∧ φ₂$. We prove $Γ ⊢ φ₁$ and $Γ ⊢ φ₂$.
From the conjunction of φ₁ and φ₂, the expected result follows.

@@

$\bullet$~Case $φ ≡ ¬ (φ₁ ∨ φ₂)$.
We prove $Γ ⊢ ¬ φ₁$ and $Γ ⊢ ¬ φ₂$.
From the conjunction of $¬ φ₁$ and $¬ φ₂$ by applying De Morgan Law
and the result follows.

\begin{equation*}
(\mathcal{D})\hspace{3mm}
\begin{bprooftree}
\AxiomC{$Γ ⊢ \fstrip₁~(¬ (φ₁ ∨ φ₂))~(\suc~n)$}
\RightLabel{by~(\ref{eq:strip-fixed})}
\UnaryInfC{$Γ ⊢ \fuh~(\fstrip₁~(¬ φ₁)~n) ∧ \fuh~(¬ φ₁ ⇒ \fstrip₁~(¬ φ₂)~n)$}
\RightLabel{∧-proj₁}
\UnaryInfC{$Γ ⊢ \fuh~(\fstrip₁~(¬ φ₁)~n)$}
\RightLabel{Lemma~\ref{lem:inv-uh-lem}}
\UnaryInfC{$Γ ⊢ \fstrip₁~(¬ φ₁)~n$}
\RightLabel{by~ind.~hyp.}
  \UnaryInfC{$Γ ⊢ ¬φ₁$}
\end{bprooftree}
\end{equation*}
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$\mathcal{D}$}
  \UnaryInfC{$Γ ⊢ ¬ φ₁$}
  \AxiomC{$Γ ⊢ \fstrip₁~(¬ (φ₁ ∨ φ₂))~(\suc~n)$}
  \RightLabel{by~(\ref{eq:strip-fixed})}
  \UnaryInfC{$Γ ⊢ \fuh~(\fstrip₁~(¬ φ₁)~n) ∧ \fuh~(¬ φ₁ ⇒ \fstrip₁~(¬ φ₂)~n)$}
  \RightLabel{∧-proj₂}
  \UnaryInfC{$Γ ⊢ \fuh~(¬ φ₁ ⇒ \fstrip₁~(¬ φ₂)~n)$}
  \RightLabel{Lemma~\ref{lem:inv-uh-lem}}
  \UnaryInfC{$Γ ⊢ ¬ φ₁ ⇒ \fstrip₁~(¬ φ₂)~n$}
  \RightLabel{⇒-elim}
  \BinaryInfC{$Γ ⊢ \fstrip₁~(¬ φ₂)~n$}
  \RightLabel{by~ind.~hyp.}
  \UnaryInfC{$Γ ⊢ ¬ φ₂$}
  \end{bprooftree}
\end{equation*}
\end{proof} % the proof continue...

The following theorem is convenient to substitute equals by equals in
a theorem. Recall the equality (≡) is symmetric and transitive as well.
We use these properties without any mention.
